Algebraic Surfaces

### 2016-05-21 project/article_revised_revised/main.tex
work work

\begin{document}
    \tableofcontents

@@

    In this section we will consider the algebraic surface

    \clearpage
    \printbibliography \printindex
\end{document}

>>>>

\begin{document}
    \maketitle

    \begin{abstract}
       In this paper we give a brief overview of some of the things one can be
       looking at when studying algebraic surfaces.  Mainly, we look at
       singularities of such surfaces, how they arise and how one can remove
       them. The images are rendered using the software \textsc{Surfer} unless
       specified otherwise.  Some auxilliary \textsc{Python} and
       \textsc{Mathematica} scripts has also been used.
    \end{abstract}

    \tableofcontents

@@

    In this section we will consider the algebraic surface given by the
    polynomial 
    \begin{equation}
        \label{eq:whitney}
        f(x, y, z) = x^2y - z^2.
    \end{equation}
    The associated variety is $\V(f(x, y, z))$. This surface is commonly known
    as the ''Whitney Umbrella''\index{Whitney umbrella} and is shown visualized in
    \cref{fig:whitney_umbrella}. As we will see shortly, when trying to blow up
    this surface in a single point, say the origin, the resulting surface does
    not seem to be any ''less singular''.

    \begin{figure}[h]
        \centering
        \includegraphics[width=0.4\linewidth]{pictures/blowup_line.png}
        \caption{The ''Whitney Umbrella'' defined by the polynomial $f(x, y, z)
        = x^2y - z^2$. This surface exhibits singularities all along the
        $y$-axis, as can easily be pointed out on the picture.}
        \label{fig:whitney_umbrella}
    \end{figure}

    \begin{example}[Blowup of the Whitney Umbrella in a single point $p$]
        If we were to proceed by trying to blow up the origin, we quickly
        encounter a problem. Following the standard procedure that we used in
        the previous example, we get the defining equations $xs = yr$, $xt =
        zr$ and $yt = zs$ with $r, s$ and $t$ our homogeneous coordinates.
        Contrary to the previous example, we here get three affine charts, one
        for each homogeneous coordinate.

        \paragraph{Case 1:}
        Setting $r = 1$ gives us $y = xs$ and $z = xt$. When plugged into
        \cref{eq:whitney} gives the equation
        \begin{equation}
            \notag
            g(x, s, t) = x^2(xs - t^2) = 0.
        \end{equation}

        \paragraph{Case 2:}
        We start by setting $s = 1$, which gives us $x = yr$ and $z = yt$.
        Eliminating $x$ and $y$ from \cref{eq:whitney} yields the equation
        \begin{equation}
            \notag
            h(y, r, t) = y^2(yr^2 - t^2) = 0.
        \end{equation}

        \paragraph{Case 3:}
        Finally, setting $t = 1$ yields $x = zr$ and $y = zs$, giving an
        equation
        \begin{equation}
            \notag
            i(z, r, s) = z^2(r^2sz - 1) = 0.
        \end{equation}
        These can be visualized, and the surfaces are shown in
        \cref{fig:blowup_line_charts}. As we see in \cref{fig:whitney_blowup} 
        the blowup fails to remove the singularities.

        \begin{figure}[h]
            \centering
            \begin{subfigure}[t]{0.3\textwidth}
                \includegraphics[width=\textwidth]{pictures/blowup_line_2.png}
                \caption{$g(x, s, t) = 0$}
            \end{subfigure}      
            ~
            \begin{subfigure}[t]{0.3\textwidth}
                \includegraphics[width=\textwidth]{pictures/blowup_line_1.png}
                \caption{$h(y, r, t) = 0$}
                \label{fig:whitney_blowup}
            \end{subfigure}      
            ~
            \begin{subfigure}[t]{0.3\textwidth}
                \includegraphics[width=\textwidth]{pictures/blowup_line_3.png}
                \caption{$i(z, r, s) = 0$}
            \end{subfigure}      
        \caption{The affine charts of the blowup of the Whitney Umbrella at the
            point $p = (0, 0, 0)$. As we can see the singularities improve in
            two of the charts, and in one case the chart is smooth. However,
            the middle surface is identical to our original surface, so the
            blow-up has failed to resolve all the singularities of the original
            variety.}
            \label{fig:blowup_line_charts}
        \end{figure}      
    \end{example}

    As seen, in this case it is not sufficient to blow up only in one single
    point. We may have to blow up the surface at a singular line, singular
    curve or in general, any other set. We make a definition of the blowup of a
    space in a line $L$, analogous to the blowup in a point,  We specialize
    this definition for $\A^3$.

    \begin{definition}[Blowup of $\A^3$ at a line $L$]
        We let the blowup-surface $\Bl_L\A^3$ be the set of all pairs
        consisting of a point $p \in \A^3$ and a point $\lambda \in \P^1$ such
        that $H_\lambda$ is the unique plane that contains both the point $p$
        and the line $L$ and write
        \begin{equation}
            \notag
            \Bl_L\A^3 = \left\{ (p, \lambda) \in \A^3 \times \P^1 \mid p, L \in H_\lambda \right\}.
        \end{equation}
    \end{definition}

    We illustrate this definition by again considering the Whitney Umbrella.
    \begin{example}[Blowup of the Whitney Umbrella at a line $L$]
        The singularities of the Whitney Umbrella lie along the $y$-axis, which
        is described by the equations $x = z = 0$. Now, our goal here is to
        find under what circumstances a plane $H_\lambda$ contains the point $p
        = (x, y, z)$ as well as the line $L$. In other words, we want both the
        $y$-axis and $p$ to lie in the plane $rx + sy + tz = 0$.

        We first find the planes that contain the $y$-axis. Choosing the vector
        $(0, 1, 0)$ as a representative for the $y$-axis we compute the cross
        product between this vector and the point $(r, s, t)$ as a
        representative for the plane.  This yields a normal vector given by
        \begin{equation}
            \notag
            \vec{n} = \begin{vmatrix}
                \hat{i} & \hat{j} & \hat{k} \\
                r & s & t \\
                0 & 1 & 0
            \end{vmatrix} = (-t, 0, r).
        \end{equation}
        A plane containing the $y$-axis is then given by the equation $-tx + 0y
        + rz = rz - tx = 0$. We can now consider the two affine charts by
        setting $t = 1$ and $s = 1$ respectively.

        \paragraph{Case 1:} 
        \label{par:case_1_}
        Setting $r = 1$ yields $z = tx$ which when plugged into $x^2y - z^2 = 0$
        gives us $x^2y - (t^2x^2) = x^2(y - t^2) = 0$. This affine chart can be
        visualized in the $(x, y, t)$-plane.

        \paragraph{Case 2:}
        \label{par:case_2_}
        Setting $t = 1$ yields $x = rz$ which when plugged into $x^2y-z^2 = 0$
        gives the equation $r^2z^2y - z^2 = r^2(z^2y - 1) = 0$. This can be
        visualized in the $(r, y, z)$-plane.

        \begin{figure}[h!]
            \centering
            \begin{subfigure}[t]{0.3\textwidth}
                \includegraphics[width=\textwidth]{pictures/line_blowup_affine_1.png}
                \caption{The surface given by the equation $x^2(y - t^2) = 0$. The
                    curve of intersection between this surface and the exceptional
                    divisor looks like it can be ''pinched'' into a double line, i.e., the
                original line of singularity.}
            \end{subfigure}
            ~
            \begin{subfigure}[t]{0.3\textwidth}
                \includegraphics[width=\textwidth]{pictures/line_blowup_affine_2.png} 
                \caption{The surface given by the equation $r^2(z^2y - 1) = 0$. The
                    intersection between this surface and the exceptional divisor can
                    also be ''fused'' together to form a double line, by laying the
                surfaces flat on top of each other.}
            \end{subfigure}
            \\
            \begin{subfigure}[t]{0.3\textwidth}
                \includegraphics[width=\textwidth]{pictures/affine_1_exceptional.pdf} 
            \end{subfigure}
            ~
            \begin{subfigure}[t]{0.3\textwidth}
                \includegraphics[width=\textwidth]{pictures/affine_2_exceptional.pdf} 
            \end{subfigure}
            \caption{The two affine charts of the variety $\V(x^2y-z^2)$. The top
                row is rendered by \textsc{Surfer}, and the bottom row rendered by
                \textsc{Mathematica.} The exceptional divisor in the top row has
                been divided out and is therefore not shown because \textsc{Surfer}
                does a poor job at rendering planes. As seen, the charts are smooth and
                exhibits no singularities. The intersection with the exceptional divisor is
            rendered in blue in the bottom row.}
        \end{figure}
    \end{example}

    While in the above example we blew up a very specific line of
    singularities, namely the $y$-axis, it can be of interest to talk a little
    bit about how to blow up general lines. It all boils down to generalizing
    the construction of the plane for a more general line $L$. Assuming $L$ is
    parametrized by $x = at, y = bt, z = ct$, with $a, b$ and $c$ non-zero, we
    can form the symmetric equations 
    \begin{equation}
        \notag
        t = \frac{x}{a} = \frac{y}{b} = \frac{z}{c}.
    \end{equation}
    The line $L$ is then completely determined by two of the three possible
    relations $xb - ya = 0$, $xc - za = 0$ and $yc - zb = 0$.  In order to find
    the respective plane, we are interested in knowing when the plane $rx + sy
    + tz = 0$ can be expressed in terms of the line given by $xb - ya = 0$ and
    $yc - zb = 0$. This can be solved as a matrix problem. Given the matrix
    \begin{equation}
        \notag
        A = \begin{bmatrix}
            b & 0 & r \\
            -a & c & s \\
            0 & -b & t
        \end{bmatrix}
    \end{equation}
    one of the columns of $A$ is linearly dependent of the two others if
    $\det(A) = 0$. Solving $\det(A) = 0$ gives the equation
    \begin{equation}
        \notag
        \det(A) = b(ct + sb) + rab = 0.
    \end{equation}
    We can now solve for either one of $r, s$ and $t$. Solving for $r$ we get
    $r = -( ct + sb ) / a$ which we can substitute back into the equation for
    our plane.  The plane is then determined by 
    \begin{equation}
        \notag
        \frac{-ct - sb}{a}x + sy + tz = 0.
    \end{equation}
    Multiplying through by $a$ we achieve
    \begin{align*}
        (-sb - ct)x + sya + tza &= -sbx - ctx + sya + tza \\
                                &= t(za - cx) + s(ya - bx) = 0,
    \end{align*}
    which gives us the restrictions that we need.
    \clearpage
    \printbibliography \printindex
\end{document}
